---
title: 'Contraceptives across space, time and their association with various social
  markers '
author: "Maria Amoah"
date: "2024-11-08"
output: html_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```

```{r, include=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(gt)
library(sf)
library(knitr)
library(ggplot2)
library(dplyr)
library(stringr)
```


Combined 2022 prescribing data comes from every month in 2022 from the [Public Health Scotalnd Open Prescribing Dataset](https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community). Data on Healthboards and population data on sex and age has been loaded in here.

```{r cars, include = FALSE}
prescribing_2022_files <-list.files(path = "data", pattern = "*.csv")


combined_2022_prescribing_data <- prescribing_2022_files %>% 
  map_dfr(~read_csv(here("data", . ))) %>% 
  clean_names()

health_boards <- read.csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()

# Eliminating the archived healthboard repeats in the datset
health_boards <- health_boards[-c (4,7,9,13),]

sex_data <- read.csv("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1/download/hb2019_pop_est_14102024.csv") %>% 
  filter(Year == "2022") %>% 
  clean_names()


```
 
*Grouping commonly prescribed contraceptive drugs into five categories:*

* Combined oral contraceptive
* Progesterone only pill
* Intrauterine contraceptives
* Contraceptive injection
* Hormone patch 

The prescribed drug brand names come from NICE who list the UK brand names of  drugs.

```{r pressurec, echo=FALSE, warning=FALSE}

birth_control_coc <- c("MICROGYON","RIGEVIDON","OVRANETTE", "CILEST", "CILIQUE", "YASMIN", "MARVELON", "GEDAREL", "ZOELY","FEMODETTE", "MILLINETTE", "SUNYA", "CIMZIT", "KATYA", "LEVEST", "LIZINNA", "FEMODENE", "LUCETTE", "BREVINOR", "OVYSMEN", "NORIMIN", "LOGYNON", "MERCILON")

birth_control_prog <- c("NORGESTON","NORIDAY","CERAZETTE", "ZELLETA", "CERELLE", "DESOGESTROL" )

iuc <- c("MIRENA", "KYLEENA", "JAYDESS", "LEVOSERT", "BENILEXA", "COPPER T380 A", "NOVAPLUS")

hormone_patch <-  "EVRA"

contraceptive_injection <- "DEPO-PROVERA"

all_contraceptives <- c(birth_control_coc, birth_control_prog, iuc, hormone_patch, contraceptive_injection)

collapsed_contraceptives <- paste(all_contraceptives, collapse = "|")

# Filtering the dataset to only include contraceptive prescriptions

all_contraceptive_data <- combined_2022_prescribing_data %>% 
  filter(str_detect(bnf_item_description,collapsed_contraceptives))

```

## Figure 1 :
Investigating the difference in contraception prescription rate across the Scottish NHS Healthboards.

The use of 2019 Spatial Data structure files is appropriate as the borders used in 2022 were determined in 2013(CITE)and have not changed
```{r, warning=FALSE, fig.cap="The distribution shows highest rates of contraception prescription in Lothian and Glasgow & Clyde,may be because these halthboards contain Scotland's two most densely popualted, big cities,  and lowest rates in Tayside. There is a general trend of higher levels in the South compared to the North." }

joined_by_hb_name <-full_join(all_contraceptive_data, health_boards, by = join_by(hbt == hb))
sex_data <- sex_data %>% 
  filter(sex != "All")
joined_with_sex <- full_join(joined_by_hb_name, sex_data, by = join_by("hbt"=="hb"))%>% 
  filter(sex != "Male") 

contraception_proportion <- joined_with_sex %>% 
  group_by(hbt) %>% 
  filter(!is.na(number_of_paid_items)) %>% 
  summarise(per_hb =sum(number_of_paid_items)/mean(all_ages))

nhs_healthboard <- st_read(here("data/NHS_healthboards_2019.shp"))

joined_with_polygon <-full_join(nhs_healthboard, contraception_proportion, by = join_by("HBCode"=="hbt")) 

contraception_distribution_hb <- joined_with_polygon %>% 
  ggplot(aes(fill = per_hb))+
   scale_fill_viridis_c(name = "Contraception prescription proportion") +
  geom_sf()+
  labs(title = "Contraception Prescribed to Women Across Scotland's Healthboards",
       subtitle = "Measured Proportionally Against Each Healthboard's Female Population") +
  theme_minimal()

contraception_distribution_hb
```



## Figure 2 : 
Observing whether or not there is a trend between age group and choice of contraceptive method

```{r, fig.cap= "Distribution across contraceptive types are similar with age 26-35 consistently being prescribed the largest number of each contraceptive type. Older age groups seem to be prescribed less progesterone only pills with the 16-25 group is slightly less than 36-45 in all except progesterone pills and 46-55 also being prescribed proprotionally less compared to the other contraceptive types."}
collapsed_coc <- paste(birth_control_coc, collapse = "|")
collapsed_p <- paste(birth_control_prog, collapse = "|")
collapsed_iuc <-  paste(iuc, collapse = "|")

# Renaming bnf_item_description column so I can group by the contraceptive categories
grouped_contraceptive_data <- joined_with_sex %>%
  mutate(bnf_item_description = if_else(str_starts(bnf_item_description, collapsed_coc), "combined_pill", bnf_item_description),
         bnf_item_description = if_else(str_starts(bnf_item_description, collapsed_p), "progesterone", bnf_item_description),
         bnf_item_description = if_else(str_starts(bnf_item_description, collapsed_iuc), "iuc", bnf_item_description),
         bnf_item_description = if_else(str_starts(bnf_item_description, "EVRA"), "hormone_patch", bnf_item_description),
         bnf_item_description = if_else(str_starts(bnf_item_description, "DEPO-PROVERA"), "injection", bnf_item_description))

#does multiplying it by number_of_paid_items this way multiply all of the ones in on bnf_item description by the same thing..

grouped_contraceptive_data <- grouped_contraceptive_data %>%
  group_by(bnf_item_description) %>%
  summarize(
    "16-25" = sum(rowSums(across(age16:age25) * number_of_paid_items), na.rm = TRUE),
    "26-35" = sum(rowSums(across(age26:age35) * number_of_paid_items), na.rm = TRUE),
    "36-45" = sum(rowSums(across(age36:age45) * number_of_paid_items), na.rm = TRUE),
    "46-55" = sum(rowSums(across(age46:age55) * number_of_paid_items), na.rm = TRUE)
  )

grouped_contraceptive_data <- grouped_contraceptive_data %>%
  pivot_longer(cols = `16-25`:`46-55`, values_to = "weighted_pop", names_to = "age")

grouped_contraceptive_data %>%
  filter(!is.na(bnf_item_description)) %>% 
  ggplot(aes(x = age, y = weighted_pop, fill = bnf_item_description)) +
  geom_col() +
  facet_wrap(~ bnf_item_description, scales = "free", labeller = labeller(bnf_item_description = c(
    "combined_pill" = "Combined Pill",
    "progesterone" = "Progesterone Only",
    "injection" = "Contraceptive Injection",
    "hormone_patch" = "Hormone Patch",
    "iuc" = "Intrauterine Contraceptive"
  ))) +
  labs(title = "Population Distribution by Age Group and Contraceptive Type",
       x = "Age Group",
       y = "Weighted Population",
       fill = "Contraceptive Type") +
  theme_minimal(base_size = 8)

```
A Limitation to this graph is that direct, person-specific comparisons between age and contraception rates were not possible so these graphs show a correlation between age distribution in healthboards and choice of contraception 

I will do it proportionally to the population of each age group? otherwise could j be showing difference in populations of different age categories. 


## Idea for Figure 3 is to make a line chart showing contraceptive rate across scotland changing over time and view it alongside factors contraception infleunces such as female employment rate, abortion rate, maternal mortality etc.?
```{r}

```

## Figure 4 - a table showing different rates of contracpetive prescription to different age groups ( expecting older population nearing menopause to have less ) and the cost per person each age group has for contraceptive prescription or something else
```{r}
#table of some sort


```

